---
title: "Climatology Downscaling"
author: "Nick Gauthier"
date: "7/14/2021"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(raster)
library(tidyverse)
library(zoo)
library(stars)
library(units)
library(tidyEOF)
library(scico)
sf_use_s2(FALSE)
```


# Setup

Setup the bounding box and a reference 5' raster for reprojection.

```{r}
bbox <- extent(c(-10, 45, 30, 50))

ref <- st_as_stars(st_bbox(bbox, crs = 4326), dx = 0.0833333)
ref_sim <- st_as_stars(st_bbox(bbox, crs = 4326), dx = 1)

coast <- rnaturalearth::ne_coastline(scale = 'medium', returnclass = 'sf') %>% 
  st_crop(ref)
countries <- rnaturalearth::ne_countries(scale = 'large', returnclass = 'sf') %>% 
  st_crop(ref)
```

Observations
```{r}
prec_obs <- list.files('/Volumes/Data/CHELSA/CHELSA_V2/climatology', full.names = TRUE, pattern = '*CHELSA_pr*') %>%
             read_stars(along = 'time') %>% 
             st_warp(ref, use_gdal = TRUE, method = 'average') %>%
             setNames('prec')
tmin_obs<- list.files('/Volumes/Data/CHELSA/CHELSA_V2/climatology', full.names = TRUE, pattern = '*CHELSA_tasmin*') %>%
             read_stars(along = 'time') %>%
             st_warp(ref, use_gdal = TRUE, method = 'average') %>%
             setNames('tmin') 
tmax_obs <- list.files('/Volumes/Data/CHELSA/CHELSA_V2/climatology', full.names = TRUE, pattern = '*CHELSA_tasmax*') %>%
             read_stars(along = 'time') %>%
             st_warp(ref, use_gdal = TRUE, method = 'average') %>%
             setNames('tmax')
srad_obs <- list.files('/Volumes/Data/CHELSA/srad', full.names = TRUE, pattern = '*CHELSA_srad*') %>%
  .[c(1,5:12, 2:4)] %>%
             read_stars(along = 'time') %>%
             st_warp(ref, use_gdal = TRUE, method = 'average') %>%
             setNames('srad')

obs <- c(prec_obs,
           tmin_obs,
           tmax_obs,
         srad_obs) %>%
  st_set_dimensions('band', values = 1:12, names = 'time') %>%
  mutate(prec = set_units(prec / 10, mm/month),
         tmin = set_units(tmin / 10, K) %>% set_units(degree_C),
         tmax = set_units(tmax / 10, K) %>% set_units(degree_C),
         srad = set_units(srad, kJ/m^2))

rm(prec_obs, tmin_obs, tmax_obs, srad_obs)
```

```{r}
prec_ccsm2 <- brick('/Volumes/Data/GCM/b40.20th.track1.1deg.005.cam2.h0.PRECT.185001-200512.nc') %>%
    .[[1465:1872]] %>% # 34 years
   rotate() %>%
  crop(bbox, snap = 'out') %>%
  stackApply(1:12, mean) %>% 
  st_as_stars() %>%
  setNames('prec2')

prec_ccsm <- brick('/Volumes/Data/GCM/pr_Amon_CCSM4_historical_r1i2p1_185001-200512.nc') %>%
    .[[1465:1872]] %>% # 34 years
   rotate() %>%
  crop(bbox, snap = 'out') %>%
  stackApply(1:12, mean) %>%
  st_as_stars() %>%
  setNames('prec')

tmin_ccsm <- brick('/Volumes/Data/GCM/tasmin_Amon_CCSM4_historical_r1i2p1_185001-200512.nc') %>%
    .[[1465:1872]] %>%
   rotate() %>%
  crop(bbox, snap = 'out') %>%
  stackApply(1:12, mean) %>%
  st_as_stars() %>%
  setNames('tmin')

tmax_ccsm <- brick('/Volumes/Data/GCM/tasmax_Amon_CCSM4_historical_r1i2p1_185001-200512.nc') %>%
    .[[1465:1872]] %>%
   rotate() %>%
  crop(bbox, snap = 'out') %>%
  stackApply(1:12, mean) %>%
  st_as_stars() %>%
  setNames('tmax')

srad_ccsm <- brick('/Volumes/Data/GCM/rsds_Amon_CCSM4_historical_r1i2p1_185001-200512.nc') %>%
    .[[1465:1872]] %>%
   rotate() %>%
  crop(bbox, snap = 'out') %>%
  stackApply(1:12, mean) %>%
  st_as_stars() %>%
  setNames('srad')

ccsm_hist <- c(prec_ccsm, tmin_ccsm, tmax_ccsm, srad_ccsm, prec_ccsm2) %>%
  st_set_dimensions('band', values = 1:12, names = 'time') %>%
  mutate(prec = set_units(prec, mm/s) %>% set_units(mm/month),
         prec2 = set_units(prec2, m/s) %>% set_units(mm/month),
        tmin = set_units(tmin, K) %>% set_units(degree_C),
         tmax = set_units(tmax, K) %>% set_units(degree_C),
        srad = set_units(srad, W/m^2))

rm(prec_ccsm, prec_ccsm2, tmin_ccsm, tmax_ccsm, srad_ccsm)
```


```{r}
ggplot() +
  geom_stars(data = obs['prec'] %>% st_set_dimensions('time', values = month.name)) +
  scale_fill_viridis_c() +
  facet_wrap(~time) +
  theme_bw() +
  labs(x = 'Longitude', y = 'Latitude') +
  coord_quickmap()
ggplot() +
  geom_stars(data = obs['tmin'] %>% st_set_dimensions('time', values = month.name)) +
  scale_fill_viridis_c(option = 'magma') +
  facet_wrap(~time) +
  theme_bw() +
  labs(x = 'Longitude', y = 'Latitude') +
  coord_quickmap()
ggplot() +
  geom_stars(data = obs['tmax'] %>% st_set_dimensions('time', values = month.name)) +
  scale_fill_viridis_c(option = 'magma') +
  facet_wrap(~time) +
  theme_bw() +
  labs(x = 'Longitude', y = 'Latitude') +
  coord_quickmap()
ggplot() +
  geom_stars(data = obs['srad'] %>% st_set_dimensions('time', values = month.name)) +
  scale_fill_viridis_c(option = 'plasma') +
  facet_wrap(~time) +
  theme_bw() +
  labs(x = 'Longitude', y = 'Latitude') +
  coord_quickmap()
```

# EOF

The spectra of each dataset look similar, so a truncation at 4 PCs seems reasonable for all.
```{r}
plot_scree(obs['prec'], k = 4, kmax = 12)
plot_scree(obs['tmin'], k = 4, kmax = 12)
plot_scree(obs['tmax'], k = 4, kmax = 12)
plot_scree(obs['srad'], k = 4, kmax = 12)
```

```{r}
plot_scree(ccsm_hist['prec'], k = 4, kmax = 12)
plot_scree(ccsm_hist['tmin'], k = 4, kmax = 12)
plot_scree(ccsm_hist['tmax'], k = 4, kmax = 12)
plot_scree(ccsm_hist['srad'], k = 4, kmax = 12)
```

```{r}
plot_scree(obs['prec'], k = 4, kmax = 12, scale = TRUE)
plot_scree(obs['tmin'], k = 4, kmax = 12, scale = TRUE)
plot_scree(obs['tmax'], k = 4, kmax = 12, scale = TRUE)
```
```{r}
plot_scree(ccsm_hist['prec'], k = 4, kmax = 12, scale = TRUE)
plot_scree(ccsm_hist['tmin'], k = 4, kmax = 12, scale = TRUE)
plot_scree(ccsm_hist['tmax'], k = 4, kmax = 12, scale = TRUE)
```


```{r}
plot_scree(ccsm_mh, k = 4, kmax = 12)
plot_scree(ccsm_hist, k = 4, kmax = 12)
```
With the scaled EOFs you could make the case for 5.
```{r}
plot_scree(era_land, k = 4, kmax = 12, scale = TRUE)
plot_scree(ccsm_mh, k = 4, kmax = 12, scale = TRUE)
plot_scree(ccsm_hist, k = 4, kmax = 12, scale = TRUE)
```


```{r}
obs_patterns_prec <- get_patterns(obs['prec'], k = 4, scale = TRUE)
obs_patterns_tmin <- get_patterns(obs['tmin'], k = 5, scale = TRUE)
obs_patterns_tmax <- get_patterns(obs['tmax'], k = 5, scale = TRUE)
```

```{r}
fit_cv <- function(dat, fun, k, obs) {
  recon <- pmap(list(dat$train_preds, dat$train_obs, dat$test), fun, k = k) %>%
    do.call('c', .)
  sqrt(mean(pull(recon - obs, 1) ^ 2, na.rm = TRUE)) %>% as.numeric()
}

library(furrr)
plan(sequential)
```

```{r}
expand_grid(kx = 1:6, ky = 1:6, kxy = 1:6) %>%
  filter(kxy <= pmin(kx, ky)) %>%
  mutate(cv = pmap_dbl(list(kx, ky, kxy), ~prep_cca(..1, ..2, 
                                                preds = ccsm_hist['srad'], 
                                                obs = obs['srad', 2:661],
                                                kfolds = 6, scale = TRUE) %>% 
                     fit_cv(fun = predict_cca, k = ..3, obs['srad', 2:661]))) %>% 
  group_by(kxy) %>% 
  filter(cv == min(cv)) %>% 
  ggplot(aes(kxy, cv)) + 
  geom_line()
```
```{r}
expand_grid(kx = 1:6, ky = 1:6, kxy = 1:6) %>%
  filter(kxy <= pmin(kx, ky)) %>%
  mutate(cv = pmap_dbl(list(kx, ky, kxy), ~prep_cca(..1, ..2, 
                                                preds = ccsm_hist['srad'], 
                                                obs = obs['srad', 2:661],
                                                kfolds = 6, scale = FALSE) %>% 
                     fit_cv(fun = predict_cca, k = ..3, obs['srad', 2:661]))) %>% 
  group_by(kxy) %>% 
  filter(cv == min(cv)) %>% 
  ggplot(aes(kxy, cv)) + 
  geom_line()
```

```{r}
expand_grid(kx = 1:6, ky = 1:6, kxy = 1:6) %>%
  filter(kxy <= pmin(kx, ky)) %>%
  mutate(cv = pmap_dbl(list(kx, ky, kxy), ~prep_cca(..1, ..2, 
                                                preds = ccsm_hist['srad'], 
                                                obs = obs['srad', 2:661],
                                                kfolds = 6, scale = FALSE, rotate = TRUE) %>% 
                     fit_cv(fun = predict_cca, k = ..3, obs['srad', 2:661]))) %>% 
  group_by(kxy) %>% 
  filter(cv == min(cv)) %>% 
  ggplot(aes(kxy, cv)) + 
  geom_line()
```

```{r}
expand_grid(kx = 1:6, ky = 1:6, kxy = 1:6) %>%
  filter(kxy <= pmin(kx, ky)) %>%
  mutate(cv = pmap_dbl(list(kx, ky, kxy), ~prep_cca(..1, ..2, 
                                                preds = ccsm_hist['srad'], 
                                                obs = obs['srad', 2:661],
                                                kfolds = 6, scale = TRUE, rotate = TRUE) %>% 
                     fit_cv(fun = predict_cca, k = ..3, obs['srad', 2:661]))) %>% 
  group_by(kxy) %>% 
  filter(cv == min(cv)) %>% 
  ggplot(aes(kxy, cv)) + 
  geom_line()
```

```{r}
expand_grid(kx = 1:8, ky = 1:8, kxy = 1:8) %>%
  filter(kxy <= pmin(kx, ky)) %>%
  mutate(cv = future_pmap_dbl(list(kx, ky, kxy), ~prep_cca(..1, ..2, 
                                                preds = ccsm_hist['prec'], 
                                                obs = obs['prec'],
                                                kfolds = 6, scale = TRUE) %>% 
                     fit_cv(fun = predict_cca, k = ..3, obs['prec'])), .progress = TRUE) %>% 
  group_by(kxy) %>% 
  filter(cv == min(cv)) %>% 
  ggplot(aes(kxy, cv)) + 
  geom_line()
```
```{r}
expand_grid(kx = 1:8, ky = 1:8, kxy = 1:8) %>%
  filter(kxy <= pmin(kx, ky)) %>%
  mutate(cv = future_pmap_dbl(list(kx, ky, kxy), ~prep_cca(..1, ..2, 
                                                preds = ccsm_hist['prec'], 
                                                obs = obs['prec'],
                                                kfolds = 6, scale = FALSE) %>% 
                     fit_cv(fun = predict_cca, k = ..3, obs['prec'])), .progress = TRUE) %>% 
  group_by(kxy) %>% 
  filter(cv == min(cv)) %>% 
  ggplot(aes(kxy, cv)) + 
  geom_line()
```
```{r}
expand_grid(kx = 1:8, ky = 1:8, kxy = 1:8) %>%
  filter(kxy <= pmin(kx, ky)) %>%
  mutate(cv = pmap_dbl(list(kx, ky, kxy), ~prep_cca(..1, ..2, 
                                                preds = ccsm_hist['prec'], 
                                                obs = obs['prec'],
                                                kfolds = 6, scale = FALSE, rotate = TRUE) %>% 
                     fit_cv(fun = predict_cca, k = ..3, obs['prec']))) %>% 
  group_by(kxy) %>% 
  filter(cv == min(cv)) %>% 
  ggplot(aes(kxy, cv)) + 
  geom_line()
```

```{r}
expand_grid(kx = 1:8, ky = 1:8, kxy = 1:8) %>%
  filter(kxy <= pmin(kx, ky)) %>%
  mutate(cv = pmap_dbl(list(kx, ky, kxy), ~prep_cca(..1, ..2, 
                                                preds = ccsm_hist['prec'], 
                                                obs = obs['prec'],
                                                kfolds = 6, scale = TRUE, rotate = TRUE) %>% 
                     fit_cv(fun = predict_cca, k = ..3, obs['prec']))) %>% 
  group_by(kxy) %>% 
  filter(cv == min(cv)) %>% 
  ggplot(aes(kxy, cv)) + 
  geom_line()
```

```{r}
expand_grid(kx = 1, ky = 1, kxy = 1) %>%
  filter(kxy <= pmin(kx, ky)) %>%
  mutate(cv = future_pmap(list(kx, ky, kxy), ~prep_cca(..1, ..2, 
                                                preds = ccsm_hist['prec'], 
                                                obs = obs['prec'],
                                                kfolds = 6, scale = TRUE, rotate = TRUE))) 

prep_cca( preds = ccsm_hist['prec'], 
                                                obs = obs['prec'],
                                                kfolds = 6, scale = TRUE, rotate = TRUE, k_preds = 2, k_obs = 2)

get_patterns(ccsm_hist['prec'], k = 1, rotate = TRUE)
get_patterns

    pca <- get_pcs(ccsm_hist['prec'], scale = TRUE)

      eofs <- pca$rotation[, 1:2, drop = FALSE] %>% # drop = FALSE preserves PC names when there's only 1 PC
      `%*%`(diag(pca$sdev,2, 2)) %>% # scale by sdev (sqrt(eigenvalues)) for more robust rotation
    `colnames<-`(paste0('PC', 1:2))
      
      varimax(eofs)
```


```{r}

expand_grid(kx = 1:8, ky = 1:8, kxy = 1:8) %>%
  filter(kxy <= pmin(kx, ky)) %>%
  mutate(cv = future_pmap_dbl(list(kx, ky, kxy), ~prep_cca(..1, ..2, 
                                                preds = ccsm_hist['tmax'], 
                                                obs = obs['tmax'],
                                                kfolds = 6, scale = TRUE) %>% 
                     fit_cv(fun = predict_cca, k = ..3, obs['tmax'])), .progress = TRUE) %>% 
  group_by(kxy) %>% 
  filter(cv == min(cv)) %>% 
  ggplot(aes(kxy, cv)) + 
  geom_line()
```
```{r}
expand_grid(kx = 1:8, ky = 1:8, kxy = 1:8) %>%
  filter(kxy <= pmin(kx, ky)) %>%
  mutate(cv = future_pmap_dbl(list(kx, ky, kxy), ~prep_cca(..1, ..2, 
                                                preds = ccsm_hist['tmin'], 
                                                obs = obs['tmin'],
                                                kfolds = 6, scale = TRUE) %>% 
                     fit_cv(fun = predict_cca, k = ..3, obs['tmin'])), .progress = TRUE) %>% 
  group_by(kxy) %>% 
  filter(cv == min(cv)) %>% 
  ggplot(aes(kxy, cv)) + 
  geom_line()
```

Do you get residual structure? ### this is wrong because it doesn't go nonneg!
```{r}
test <- get_patterns(obs['tmax'] -reconstruct_field(get_patterns(obs['tmax'], scale = TRUE, k  = 1)), k = 5, scale = TRUE)
test7 <- get_patterns(obs['tmax'] -reconstruct_field(get_patterns(obs['tmax'], scale = TRUE, k  = 1), nonneg = FALSE), k = 5, scale = TRUE)

test2 <- get_patterns(obs['tmin'] -reconstruct_field(get_patterns(obs['tmin'], scale = TRUE, k  = 1)), k = 5, scale = TRUE)
test8 <- get_patterns(obs['tmin'] -reconstruct_field(get_patterns(obs['tmin'], scale = TRUE, k  = 1), nonneg = FALSE), k = 5, scale = TRUE)

test3 <- get_patterns(obs['srad', 2:661] - reconstruct_field(get_patterns(obs['srad',2:661], scale = TRUE, k  = 1)), k = 5, scale = TRUE)
test9 <- get_patterns(obs['srad', 2:661] - reconstruct_field(get_patterns(obs['srad', 2:661], scale = TRUE, k  = 1), nonneg = FALSE), k = 5, scale = TRUE)

plot_eofs(test)
plot_eofs(test7)

plot_eofs(test2)
plot_eofs(test8)

plot_eofs(test3)
plot_eofs(test9)
```


```{r}
plot(reconstruct_field(get_patterns(obs['srad',2:661], scale = TRUE, k  = 1)))
plot(reconstruct_field(get_patterns(obs['srad',2:661], scale = FALSE, k  = 1)))

plot_eofs(get_patterns(obs['srad',2:661], scale = TRUE, k  = 1))
plot_eofs(get_patterns(obs['srad',2:661], scale = FALSE, k  = 1))

plot(obs['srad', 2:661] - reconstruct_field(get_patterns(obs['srad',2:661], scale = TRUE, k  = 1)))
```

```{r}
plot_eofs(obs_patterns_prec)
plot_eofs(obs_patterns_tmin)
plot_eofs(obs_patterns_tmax)
```
```{r}
plot_amps(obs_patterns_prec)
plot_amps(obs_patterns_tmin)
plot_amps(obs_patterns_tmax)
```


```{r}
hist_patterns_prec <- get_patterns(ccsm_hist['prec'], k = 4, scale = TRUE)
hist_patterns_tmin <- get_patterns(ccsm_hist['tmin'], k = 5, scale = TRUE)
hist_patterns_tmax <- get_patterns(ccsm_hist['tmax'], k = 5, scale = TRUE)
```


```{r}
plot_eofs(hist_patterns_prec)
plot_eofs(hist_patterns_tmin)
plot_eofs(hist_patterns_tmax)
```
```{r}
plot_amps(obs_patterns_prec)
plot_amps(obs_patterns_tmin)
plot_amps(obs_patterns_tmax)
plot_amps(hist_patterns_prec)
plot_amps(hist_patterns_tmin)
plot_amps(hist_patterns_tmax)
```

```{r}
mh_prec <- brick('/Volumes/Data/PMIP3//MH/pr/pr_Aclim_CCSM4_midHolocene_r1i1p1_100001-130012-clim.nc', var = 'pr') %>%
  raster::rotate() %>%
  raster::crop(bbox, snap = 'out') %>%
  st_as_stars() %>%
  setNames('prec') 

mh_tmin <- brick('/Volumes/Data/PMIP3/MH/tasmin/tasmin_Aclim_CCSM4_midHolocene_r1i1p1_100001-130012-clim.nc', var = 'tasmin') %>%
  raster::rotate() %>%
  raster::crop(bbox, snap = 'out') %>%
  st_as_stars() %>%
  setNames('tmin') 

mh_tmax <- brick('/Volumes/Data/PMIP3/MH/tasmax/tasmax_Aclim_CCSM4_midHolocene_r1i1p1_100001-130012-clim.nc', var = 'tasmax') %>%
  raster::rotate() %>%
  raster::crop(bbox, snap = 'out') %>%
  st_as_stars() %>%
  setNames('tmax') 

ccsm_mh <- c(mh_prec, mh_tmin, mh_tmax) %>%
  st_set_dimensions('band', values = 1:12, names = 'time') %>%
  mutate(prec = set_units(prec, mm/s) %>% set_units(mm/month),
        tmin = set_units(tmin, K) %>% set_units(degree_C),
         tmax = set_units(tmax, K) %>% set_units(degree_C))
```


# CV
So set up a cross validation scheme to compare the scaled/unscaled, rotated/unrotated eofs, and number of EOFs.

# Anomalies

```{r}
ggplot() +
  geom_stars(data = (ccsm_mh['prec'] - ccsm_hist['prec']) %>% st_set_dimensions('time', values = month.name)) +
  scale_fill_distiller(palette = 'PuOr', limits = c(-28, 28), direction = 1) +
  facet_wrap(~time) +
  theme_bw() +
  labs(x = 'Longitude', y = 'Latitude', title = 'Monthly mean preciptiation anomaly', subtitle = 'Mid Holocene to Present') +
  coord_quickmap()

ggplot() +
  geom_stars(data = (ccsm_mh['prec'] - ccsm_hist['prec2']) %>% st_set_dimensions('time', values = month.name)) +
  scale_fill_distiller(palette = 'PuOr', limits = c(-28, 28), direction = 1) +
  facet_wrap(~time) +
  theme_bw() +
  labs(x = 'Longitude', y = 'Latitude', title = 'Monthly mean preciptiation anomaly', subtitle = 'Mid Holocene to Present') +
  coord_quickmap()

ggplot() +
  geom_stars(data = (ccsm_mh['tmax'] - ccsm_hist['tmax']) %>% st_set_dimensions('time', values = month.name)) +
  scale_fill_distiller(palette = 'RdBu', limits = c(-4.8, 4.8)) +
  facet_wrap(~time) +
  theme_bw() +
  labs(x = 'Longitude', y = 'Latitude', title = 'Monthly mean maximum temperature anomaly', subtitle = 'Mid Holocene to Present') +
  coord_quickmap()

ggplot() +
  geom_stars(data = (ccsm_mh['tmin'] - ccsm_hist['tmin']) %>% st_set_dimensions('time', values = month.name)) +
  scale_fill_distiller(palette = 'RdBu', limits = c(-4.1, 4.1)) +
  facet_wrap(~time) +
  theme_bw() +
  labs(x = 'Longitude', y = 'Latitude', title = 'Monthly mean minimum temperature anomaly', subtitle = 'Mid Holocene to Present') +
  coord_quickmap()
```
```{r}
preind <- brick('~/Downloads/pr_Aclim_CCSM4_piControl_r1i1p1_025001-130012-clim.nc', var = 'pr') %>%
  raster::rotate() %>%
  raster::crop(bbox, snap = 'out') %>%
  st_as_stars() %>%
   #   st_warp(ref_sim, use_gdal = TRUE, method = 'bilinear') %>%
  st_set_dimensions('band', values = 1:12, names = 'time') %>%
  setNames('prec')%>%
  mutate(prec = units::set_units(prec, mm/s)) %>%
  mutate(prec = units::set_units(prec, mm/month))

mri <- brick('/Volumes/Data/PMIP3/MH/pr/pr_Aclim_MRI-CGCM3_midHolocene_r1i1p1_195101-205012-clim.nc', var = 'pr') %>%
  raster::rotate() %>%
  raster::crop(bbox, snap = 'out') %>%
  st_as_stars() %>%
   #   st_warp(ref_sim, use_gdal = TRUE, method = 'bilinear') %>%
  st_set_dimensions('band', values = 1:12, names = 'time') %>%
  setNames('prec')%>%
  mutate(prec = units::set_units(prec, mm/s)) %>%
  mutate(prec = units::set_units(prec, mm/month))

giss <- brick('/Volumes/Data/PMIP3/MH/pr/pr_Aclim_GISS-E2-R_midHolocene_r1i1p1_250001-259912-clim.nc', var = 'pr') %>%
  raster::rotate() %>%
  raster::crop(bbox, snap = 'out') %>%
  st_as_stars() %>%
   #   st_warp(ref_sim, use_gdal = TRUE, method = 'bilinear') %>%
  st_set_dimensions('band', values = 1:12, names = 'time') %>%
  setNames('prec')%>%
  mutate(prec = units::set_units(prec, mm/s)) %>%
  mutate(prec = units::set_units(prec, mm/month))
```

```{r}
plot_amps(hist_patterns_prec)
plot_amps(obs_patterns_prec)
plot_amps(get_patterns(obs['prec'], scale = FALSE, rotate = TRUE))
plot_amps(get_patterns(ccsm_mh['prec'], scale = FALSE, rotate = TRUE))
plot_amps(get_patterns(ccsm_hist['prec'], scale = FALSE, rotate = TRUE))
plot_amps(get_patterns(ccsm_hist['prec2'], scale = FALSE, rotate = TRUE))
plot_amps(get_patterns(preind, scale = FALSE, rotate = TRUE))

```

```{r}

a <- bind_rows(list(obs = get_patterns(obs['prec'])$amplitudes,
          prec_hist = get_patterns(ccsm_hist['prec'])$amplitudes,
          prec2_hist = get_patterns(ccsm_hist['prec2'])$amplitudes,
          mh = get_patterns(ccsm_mh['prec'])$amplitudes,
          preind = get_patterns(preind)$amplitudes,
          mri = get_patterns(mri)$amplitudes,
          giss = get_patterns(giss)$amplitudes), .id = 'type') %>%
  ggplot(aes(time, PC1, color = type)) +
    scale_color_brewer(palette = 'Spectral') +
  geom_line(size = 1.2, alpha = .8) +
  ggtitle('original')


b <- bind_rows(list(obs = get_patterns(obs['prec'], rotate = TRUE)$amplitudes,
          prec_hist = get_patterns(ccsm_hist['prec'], rotate = TRUE)$amplitudes,
          prec2_hist = get_patterns(ccsm_hist['prec2'], rotate = TRUE)$amplitudes,
          mh = get_patterns(ccsm_mh['prec'], rotate = TRUE)$amplitudes,
          preind = get_patterns(preind, rotate = TRUE)$amplitudes,
          mri = get_patterns(mri, rotate = TRUE)$amplitudes,
          giss = get_patterns(giss, rotate = TRUE)$amplitudes), .id = 'type') %>%
  ggplot(aes(time, PC1, color = type)) +
    scale_color_brewer(palette = 'Spectral') +
  geom_line(size = 1.2, alpha = .8) +
  ggtitle('rotate')

c <- bind_rows(list(obs = get_patterns(obs['prec'], scale = TRUE, rotate = TRUE)$amplitudes,
          prec_hist = get_patterns(ccsm_hist['prec'], scale = TRUE, rotate = TRUE)$amplitudes,
          prec2_hist = get_patterns(ccsm_hist['prec2'], scale = TRUE, rotate = TRUE)$amplitudes,
          mh = get_patterns(ccsm_mh['prec'], scale = TRUE, rotate = TRUE)$amplitudes,
          preind = get_patterns(preind, scale = TRUE, rotate = TRUE)$amplitudes,
          mri = get_patterns(mri, scale = TRUE, rotate = TRUE)$amplitudes,
          giss = get_patterns(giss, scale = TRUE, rotate = TRUE)$amplitudes), .id = 'type') %>%
  ggplot(aes(time, PC1, color = type)) +
    scale_color_brewer(palette = 'Spectral') +
  geom_line(size = 1.2, alpha = .8) +
  ggtitle('rotate_scale')


d <- bind_rows(list(obs = get_patterns(obs['prec'], scale = TRUE)$amplitudes,
          prec_hist = get_patterns(ccsm_hist['prec'], scale = TRUE)$amplitudes,
          prec2_hist = get_patterns(ccsm_hist['prec2'], scale = TRUE)$amplitudes,
          mh = get_patterns(ccsm_mh['prec'], scale = TRUE)$amplitudes,
          preind = get_patterns(preind, scale = TRUE)$amplitudes,
          mri = get_patterns(mri, scale = TRUE)$amplitudes,
          giss = get_patterns(giss, scale = TRUE)$amplitudes), .id = 'type') %>%
  ggplot(aes(time, PC1, color = type)) +
    scale_color_brewer(palette = 'Spectral') +
  geom_line(size = 1.2, alpha = .8) +
  ggtitle('scale')

library(patchwork)
a + b + c +d +plot_layout(guides = 'collect')
```


```{r}
plot_eofs(get_patterns(obs['prec']))
plot_eofs(get_patterns(ccsm_hist['prec']))

plot_eofs(get_patterns(obs['prec'], scale = TRUE))
plot_eofs(get_patterns(ccsm_hist['prec'], scale = TRUE))

plot_eofs(get_patterns(obs['prec'], scale = TRUE, rotate = TRUE))
plot_eofs(get_patterns(ccsm_hist['prec'], scale = TRUE, rotate = TRUE))
```

# Downscaling


```{r}
prec_recon <- predict_cca(hist_patterns_prec, obs_patterns_prec, ccsm_mh['prec'], k = 4)
tmin_recon <- predict_cca(hist_patterns_tmin, obs_patterns_tmin, ccsm_mh['tmin'], k = 4)
tmax_recon <- predict_cca(hist_patterns_tmax, obs_patterns_tmax, ccsm_mh['tmax'], k = 4)
```

```{r}
ggplot() +
  geom_stars(data = t1) +
  facet_wrap(~time) +
  scale_fill_viridis_c(na.value = NA) +
  coord_quickmap() +
  theme_bw()
ggplot() +
  geom_stars(data = t2) +
  facet_wrap(~time) +
  scale_fill_viridis_c(na.value = NA) +
  coord_quickmap() +
  theme_bw()
```

Ok so as a first look there is a good deal of disagreement between the CCSM anomalies and the downscaled ones.

```{r}
delta <- function(pred, obs, newdata = NULL, k = NULL) { # k is just a placeholder
  if(is.null(newdata)) newdata <- pred
  pred_clim <- get_climatology(pred)
  obs_clim <- get_climatology(obs)

  (newdata / pred_clim['mean']) %>%
    st_warp(slice(obs, 'time', 1), use_gdal = TRUE, method = 'bilinear') %>%
    setNames(names(newdata)) %>%
    mutate(across(everything(), ~units::set_units(.x, units(newdata[[1]]), mode = 'standard'))) %>%
    st_set_dimensions('band', values = st_get_dimension_values(newdata, 'time'), names = 'time') %>%
    `*`(obs_clim['mean'])
}

delta_add <- function(pred, obs, newdata = NULL, k = NULL) { # k is just a placeholder
  if(is.null(newdata)) newdata <- pred
  pred_clim <- get_climatology(pred)
  obs_clim <- get_climatology(obs)

  # simplify units here!
  (units::drop_units(newdata) - pred_clim['mean']) %>%
    st_warp(slice(obs, 'time', 1), use_gdal = TRUE, method = 'bilinear') %>%
    setNames(names(newdata)) %>%
    mutate(across(everything(), ~units::set_units(.x, units(newdata[[1]]), mode = 'standard'))) %>%
    st_set_dimensions('band', values = st_get_dimension_values(newdata, 'time'), names = 'time') %>%
    units::drop_units() %>%
    `+`(obs_clim['mean']) %>%
    mutate(prec = units::set_units(if_else(prec < 0, 0, prec), mm/month))
}

delta1 <- delta(ccsm_hist, era_land, newdata = ccsm_mh)
delta2 <- delta_add(ccsm_hist, era_land, newdata = ccsm_mh)
```

```{r}
gam <- read_stars('~/Downloads/ccsm4_prc_mh6k.tif') %>%
  st_warp(era_land, method = 'bilinear', use_gdal = TRUE) %>%
  st_set_dimensions('band', names = 'time')
plot(gam)

trace_mh <- list.files('/Volumes/Data/CHELSA/chelsa_trace/pr', pattern = '*_-40_*', full.names = TRUE)[c(1,5:12,2:4)] %>%
  read_stars(along = 'time') %>%
  st_warp(era_land, use_gdal = TRUE, method = 'bilinear') %>%
    st_set_dimensions('band', names = 'time')
trace_mh2 <- list.files('/Volumes/Data/CHELSA/chelsa_trace/pr', pattern = '*_-45_*', full.names = TRUE)[c(1,5:12,2:4)] %>%
  read_stars(along = 'time') %>%
  st_warp(era_land, use_gdal = TRUE, method = 'bilinear') %>%
    st_set_dimensions('band', names = 'time')

trace_present <- list.files('/Volumes/Data/CHELSA/chelsa_trace/pr', pattern = '*_20_*', full.names = TRUE)[c(1,5:12,2:4)] %>%
  read_stars(along = 'time') %>%
  st_warp(era_land, use_gdal = TRUE, method = 'bilinear') %>%
    st_set_dimensions('band', names = 'time')
```
```{r}
plot(trace_mh - trace_mh2)
```


```{r}
ggplot() +
  geom_stars(data = t1 - era_land) +
  facet_wrap(~time) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-27, 27), direction = 1) +
  geom_sf(data= coast) +
  theme_void()
ggplot() +
  geom_stars(data = st_crop(ccsm_mh - ccsm_hist, countries)) +
  facet_wrap(~time) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-27, 27), direction = 1) +
  geom_sf(data= coast) +
  theme_void()
ggplot() +
  geom_stars(data = st_crop(delta1 - era_land, countries)) +
  facet_wrap(~time) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-60, 60), direction = 1) +
  geom_sf(data= coast) +
  theme_void()
ggplot() +
  geom_stars(data = st_crop(delta2 - era_land, countries)) +
  facet_wrap(~time) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-60, 60), direction = 1) +
  geom_sf(data= coast) +
  theme_void()
ggplot() +
  geom_stars(data = st_crop(trace_mh - trace_present, countries)) +
  facet_wrap(~time) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-60, 60), direction = 1) +
  geom_sf(data= coast) +
  theme_void()
```
Hmmm, it looks like there could be a date mismatch somewhere here? for example the november and december
So this *could* be the calendar shift, that in theory is not captured by the delta methods but is captured by the EOF method.
Very weird, sometimes the sign of the delta method is totally different from the raw anomalies . . .that shouldn't happen right?

Try plotting annual and seasonal anomalies to compare to previous work.
```{r}
a <- ggplot() +
  geom_stars(data = (st_apply(t1, 1:2, sum) - st_apply(era_land, 1:2, sum)) / 30) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-5, 5), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

b <- ggplot() +
  geom_stars(data = (st_apply(t2, 1:2, sum) - st_apply(era_land, 1:2, sum)) / 30) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-5, 5), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

c <- ggplot() +
  geom_stars(data = (st_apply(ccsm_mh, 1:2, sum) - st_apply(ccsm_hist, 1:2, sum)) / 30) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-5, 5), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

d <- ggplot() +
  geom_stars(data = (st_apply(delta1, 1:2, sum) - st_apply(era_land, 1:2, sum)) / 30) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-5, 5), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

e <- ggplot() +
  geom_stars(data = (st_apply(delta2, 1:2, sum) - st_apply(era_land, 1:2, sum)) / 30) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-5, 5), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

f <- ggplot() +
  geom_stars(data = (st_apply(trace_mh, 1:2, sum) - st_apply(trace_present, 1:2, sum)) / 30) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-5, 5), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

library(patchwork)
a + b + c + d + e + f + plot_layout(guides = 'collect')
```

```{r}
g <- ggplot() +
  geom_stars(data = (st_apply(t1[,,,c(1,2,12)], 1:2, sum) - st_apply(era_land[,,,c(1,2,12)], 1:2, sum))) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-100, 100), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

h <- ggplot() +
  geom_stars(data = (st_apply(t2[,,,c(1,2,12)], 1:2, sum) - st_apply(era_land[,,,c(1,2,12)], 1:2, sum))) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-100, 100), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

i <- ggplot() +
  geom_stars(data = (st_apply(ccsm_mh[,,,c(1,2,12)], 1:2, sum) - st_apply(ccsm_hist[,,,c(1,2,12)], 1:2, sum))) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-100, 100), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

j <- ggplot() +
  geom_stars(data = (st_apply(delta1[,,,c(1,2,12)], 1:2, sum) - st_apply(era_land[,,,c(1,2,12)], 1:2, sum))) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-100, 100), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

k <- ggplot() +
  geom_stars(data = (st_apply(delta2[,,,c(1,2,12)], 1:2, sum) - st_apply(era_land[,,,c(1,2,12)], 1:2, sum))) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-100, 100), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

l <- ggplot() +
  geom_stars(data = (st_apply(trace_mh[,,,c(1,2,12)], 1:2, sum) - st_apply(trace_present[,,,c(1,2,12)], 1:2, sum))) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-100, 100), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

g + h + i + j + k + l + plot_layout(guides = 'collect')
```

```{r}
m <- ggplot() +
  geom_stars(data = (st_apply(t1[,,,6:8], 1:2, mean) - st_apply(era_land[,,,6:8], 1:2, mean))) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-50, 50), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

n <- ggplot() +
  geom_stars(data = (st_apply(t2[,,,6:8], 1:2, mean) - st_apply(era_land[,,,6:8], 1:2, mean))) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-50, 50), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

o <- ggplot() +
  geom_stars(data = (st_apply(ccsm_mh[,,,6:8], 1:2, mean) - st_apply(ccsm_hist[,,,6:8], 1:2, mean))) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-50, 50), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

p <- ggplot() +
  geom_stars(data = (st_apply(delta1[,,,6:8], 1:2, mean) - st_apply(era_land[,,,6:8], 1:2, mean))) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-50, 50), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

q <- ggplot() +
  geom_stars(data = (st_apply(delta2[,,,6:8], 1:2, mean) - st_apply(era_land[,,,6:8], 1:2, mean))) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-50, 50), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

r <- ggplot() +
  geom_stars(data = (st_apply(trace_mh[,,,6:8], 1:2, mean) - st_apply(trace_present[,,,6:8], 1:2, mean))) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-50, 50), direction = 1) +
  geom_sf(data= coast) +
  theme_void()

m + n + o + p + q + r + plot_layout(guides = 'collect')
```

