---
title: "Eco-cultural niches and the spread of farming"
author: "Nick Gauthier"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    fig_caption: yes
    highlight: zenburn
    keep_md: yes
    theme: flatly
    toc: yes
---
## Introduction
This is a research compendium to gather a handful of analyses I've been exploring how various eco-cultural niches have influenced the westward spread of farming from the Eastern Mediterranean. After importing the database of archaeological sites and radiocarbon dates, we first compare the dates to simple geographic distances from a selected "origin" site. Then, we incorporate information from crop models, driven by downscaled paleoclimate simulations, to see if crop productivity influenced the spread of farming. Throughout, we use smooth additive quantile regression models for statistical inference, allowing us to build flexible models that target the earliest sites in the database.

## Setup
Load necessary packages. These are all available on CRAN.
```{r message = F, warning = F}
library(raster) # for processing of raster maps
library(tidyverse) # for data cleaning and plotting
library(mgcv) # for fitting generalized additive model 
library(gdistance) # for least cost path calculations
library(viridis) # for pretty colors
library(qgam) # for quantile regression
```

Load (and if necessary download) additional packages from github.
```{r message = F, warning = F}
# if gganimate or schoenberg aren't already installed, uncomment and run the next lines:
# devtools::install_github('dgrtwo/gganimate')
# devtools::install_github("gavinsimpson/schoenberg")
library(gganimate) # for animated plots
library(schoenberg) # for plotting fitted GAMs
```

## Archaeological Data
Import the database of 14C dates from Neolithic sites.
```{r message = F}
dates <- read_csv('data/cs_dates_4_grass.csv') %>%
  filter(Long != 0) %>% # get rid of erroneous entry
  select(Site_Name, Long, Lat, Cal.BC, Cal.BC.SD) # select just the variables we'll need
```

Plotting out these data reveals a clear temporal discontinuity, with clusters of very early sites located around the "core" areas of domestication and later sites beyond. This pattern is consistent with the hypothesized spread of *farming* within the core Eastern Mediterranean zone, and the spread of *farmers* into Europe.
```{r echo = F}
load('data//med_region_gg.rda')

ggplot(dates, aes(Long, Lat)) +
  geom_point(aes(color = Cal.BC)) +
  geom_path(data = med_region_gg, aes(long,lat, group = group)) +
  scale_color_distiller(palette = 'RdYlBu') +
  coord_quickmap(xlim = c(-9.9, 45), ylim = c(29.8, 50)) +
  labs(title = 'Neolithic spread', subtitle = 'Median calibrated data of initial site occupation') +
  theme_void()
```
## Cost-Distance Modeling
### Geographical Predictors
First, we compare the initial site dates in our database to simple geographic variables (mainly distance from an "origin" site). This allows us to establish baseline estimates of the spreading dynamic, to compare with crop-model simulations later.

Define our study area as the supra-Mediterranean region.
```{r}
bbox <- extent(-9.9, 45, 30, 50) # switch 30 to 29.8 to keep extra site in
```

Import map of distances from the coast, and use to make masks to distinguish land and water areas ([source](http://worldgrids.org/doku.php/wiki:dicgsh1)).
```{r}
distance_coast <- raster('data/DICGSH1a.tif') %>% # import the map
  crop(bbox) %>% # crop to the study area
  reclassify(c(-Inf, 0, 0)) # set land areas to 0

land_sea <-  distance_coast %>%
  reclassify(c(-Inf, Inf, 1)) # set ALL cells in the study area to 1

land_mask <- distance_coast %>%
  reclassify(c(0, Inf, 0, -Inf, 0, 1)) # set water cells to 0, all else to 1
```

```{r echo = F}
land_mask %>%
  as.data.frame(xy = T, na.rm = T) %>%
  mutate(layer = if_else(layer == 0, 'Sea', 'Land')) %>%
  ggplot(aes(x, y, fill = layer)) +
    geom_raster() +
    ggtitle('Land-sea mask', 'Source: Worldgrids.org') +
    coord_quickmap() +
    theme_void()
```

Let's use the **gdistance** package to calculate how far all the sites in our database are from Jericho (which we use as the "first" site). In reality, Cayonu is the earliest site in our database, and there is a case to be made for using both sites as dual origin points (explored below), but for now we start with Jericho to keep things simple.

Find the location of Jericho to use as a starting point. 
```{r}
jericho <- dates %>%
  filter(Site_Name == 'Jericho') %>% # get the entry for Jericho
  select(Long:Lat) %>% # pull the location information
  as.matrix # convert to a matrix
```

Now generate a few simple distance matrices using 3 different geographical models:
1. Great-circle distance from Jericho.
2. Least cost distance from Jericho over land.
3. Least cost distance from Jericho over land and water.

First we calculate the simple great circle distance from all sites to Jericho, as a baseline metric of geographic diffision. 
```{r}
gc_dist <- dates %>% # start with the 14C databse
  select(Long:Lat) %>% # pull out the locations
  as.matrix %>% # convert into matrix format
  pointDistance(jericho, ., lonlat = T) # calculate great circle distances
```


Next we use **gdistance** to incorporate the different impacts of traveling over land or sea, exploring a range of assumptions for how easy ocean travel was to early farmers. **gdistance** is similar to other cost-distance modeling tools in GRASS, save for that instead of using friction maps to estimate the "cost" or "resistance" to movement it uses maps of "conductance" (1/cost). This leads to much improved computational performance. 

Transition matrices are the core data type used in **gdistance**, think of them as simple spatial networks where the edge weights represent the conductance, or more simply the ease of travel, between two neighboring grid points. For simplicity, we set the conductance between all land cells to 1. Multiplying these conductances by the distance between the relevant cells results in a "cost" of travel equal to the distance between cells.

Generate a simple conductance matrix, where all land-land connections are set to 1 and all land-sea or sea-sea connections are set to 0.
```{r}
con_land <- transition(land_mask, min, 16) # 16 = knight's case for neighborhoods
```

Also make one where all connections, land-land and land-sea, are 1. We'll use this as an intermeidate step later on
```{r}
con_landsea <- transition(land_sea, max, 16) 
```


Both the great-circle distances and land-only distances are unrealistic, as they assume the ocean either has no or infinite resistance to travel. Let's calculate a continuous decay function as distance from the coast increases, so that cells close to the shore have conductances near 1. Define a function to do this after Van Etten and Hijmans 2011.
```{r}
decay <- function(x, p){
	tm <- transitionMatrix(x)
	tm@x[tm@x>0.1] <- (2^-(tm@x[tm@x>0.1]/p))
	tm@x <- tm@x/max(tm@x)
	transitionMatrix(x) <- tm
	return(x)
}
```

The parameter **p** in the above function is the conductance half value of distance, measured in kilometers. Varying this parameter changes how far out to sea someone can easily travel. Let's calculate transition matrices for a range of such values.
```{r}
coastal_decay <- distance_coast %>% # start with the distance to the coast map
  transition(mean, directions = 16) # turn into a transition matrix

coastal_decay <- seq(10, 90, 10) %>% # different values for p
  map(~ decay(coastal_decay, .)) # apply the decay function to the transition matrix for each value of p
```

```{r echo = F}
coastal_decay %>%
  map(raster) %>%
  brick %>%
  `names<-`(factor(paste0('p', seq(10, 90, 10)))) %>% 
  as.data.frame(xy = T, na.rm = T) %>%
  gather(key = decay_constant, value, 3:11) %>%
  ggplot(aes(x, y)) +
    geom_raster(aes(fill = value)) +
    facet_wrap(~decay_constant) +
    scale_fill_distiller(palette = 'YlGnBu', direction = 1, name = 'Conductance') +
    ggtitle('Distance decay for ocean travel', 'Under varying conductance half value constants') +
    coord_quickmap() +
    theme_void()
```

As noted above, these raw conductance values must be corrected to accound for intercell distances. Calculate this correction matrix, which we can later multiply by the transition matrices to get the corrected values.
```{r}
correction_lcp <- geoCorrection(con_landsea, type = 'c', multpl = T)
```


Calculate the least cost distances from Jericho for each of the conductance layers, correcting for intercell distances.
```{r}
distances <- c(con_landsea, # put everything together in a list
               con_land,
               map(coastal_decay, ~. + con_land)) %>% # combine land and sea conductances
  map(~ . * correction_lcp) %>% # apply the geocorrection
  map(accCost, fromCoords = jericho) %>% # calculate the accumulated cost distances
  brick %>% # convert to a raster brick
  `names<-`(c('Euclidean distance', 'Land only', paste('p', seq(10, 90, 10))))
```

```{r echo = F}
distances %>%
  mask(land_mask, maskvalue = 0) %>% # mask the distances to land areas
  as.data.frame(xy = T, na.rm =T) %>%
  gather(type, cost, 3:13) %>%
  #filter(cost < 1e10) %>% # remove outliers (inaccessible islands)
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = cost)) +
  facet_wrap(~type, ncol = 3) +
  scale_fill_viridis(direction = -1) +
  ggtitle('Least-cost distances from Jericho') +
  coord_quickmap() +
  theme_void()
```

Sample these maps at the location of the sites, and add the results to the sites data.
```{r}
costs <- dates %>%
  select(Long:Lat) %>%
  as.matrix %>%
  raster::extract(distances, ., df = T) %>%
  cbind(dates, great_circle = gc_dist, .) %>%
  select(-ID) %>%
  gather(type, cost, 6:17) %>%
  nest(-type) %>%
  mutate(data = map(data, ~filter(., is.finite(cost)))) %>%
  mutate(data = map(data, ~filter(., cost < 1e10))) # remove outliers
```

### Statistical Inference
Now that we've generated a bunch of estimates of the "costs" of moving from Jericho to all the other sites, let's see how well these cost maps can predict the actual pattern of arrival times encoded in the 14C dates. Because we have at least two phases of spread (cultural and demic) a single linear model would be insufficient to capture the dynamics we are interested in, as that approach assumes a constant rate of spread. Instead we fit splines to the data, allowing the spread rates to be flexibly be estimated from the data themselves and vary in time. Furthermore, instead of regressing against the mean, we use a quantile regression approach to model the top 15th percentile of the earliest sites. This approach ensures that the signal of the initial, "colonizer" sites in a given area is not swamped out by later foundations from after the initial introduction of farming.

First define two helper functions. The *quantile_gam()* function is where the GAM actual get's fit. We write and save it here, so we can easily apply it to all the different datasets automatically.
```{r}
quantile_gam <- function(df){
  qgam(list(
    Cal.BC ~ s(cost, bs = 'cr'), 
           ~ s(cost, bs = 'cr')), 
    data = df,
    qu = 0.8)
}
```

The next function allows us to extract the splines and confidence intervals from models generated from the above function.
```{r}
make_preds <- function(x){
  predict(x, type = 'response', se = T) %>%
  map(~.[,1]) %>%
  bind_cols %>%
  mutate(upper = fit + 2 * se.fit,
         lower = fit - 2 * se.fit)
}
```

Now fit a quantile GAM to each of the different cost distance datasets, extracting the AIC and predictions of each model.
```{r message = F}
fits <- costs %>%
  mutate(mod = map(data, quantile_gam),
         aic = map_dbl(mod, AIC),
         reml = map_dbl(mod, 'gcv.ubre'),
         preds = map(mod, make_preds))
fits$mod
  walk(fits$mod, ~check(.$calibr, 2))
```

Let's look at the results, specifically comparing the AICs across each model. At first glance, the land only model has the best fit. But wait! The land only model was fit on ~60 fewer sites, because by definition it cannot account for arrival times at island sites. With a smaller sample size, the AIC must be lower, so we should discount this model. So here the suggestion is that the land-sea model with a small (10) conductance half value parameter fits the data best. This is a consistent with Neolithic farmers capable of traveling over the sea, but only very short distances away from the coast.
```{r}
fits %>%
  arrange(aic) %>%
  select(type, aic) %>% 
  ggplot(aes(as.character(type), aic)) +
  geom_point()
fits %>%
  arrange(reml) %>%
  select(type, reml)
```
Just to make sure, let's rerun the above analysis, limiting the the regressions only to sites on land (so that the AICs across the land and other models will be comparable). Of course, this isn't ideal either because part of the benefit of including the land-sea models is that they can account for arrival times on islands. Now we see the best fitting model has p = 70, with p = 10 coming in second. Let's stick with the p = 10 model, as it 
```{r echo = F}
dates %>%
  select(Long:Lat) %>%
  as.matrix %>%
  raster::extract(distances, ., df = T) %>%
  cbind(dates, great_circle = gc_dist, .) %>%
  select(-ID) %>%
  drop_na() %>%
  filter(is.finite(Land.only)) %>%
  gather(type, cost, 6:17) %>%
  nest(-type) %>% 
  mutate(mod = map(data, quantile_gam),
         aic = map_dbl(mod, AIC)) %>%
  arrange(aic) %>%
  select(type, aic) %>%
  .$aic %>%
  plot
```

```{r}
plot_results <- function(x){
  print(ggplot(x, aes(cost, Cal.BC)) +
  geom_pointrange(aes(ymin = Cal.BC - Cal.BC.SD, 
                      ymax = Cal.BC + Cal.BC.SD,  
                      alpha = (1 / Cal.BC.SD) ^ 2), fatten = 2) +
  geom_line(aes(y = fit), color = 'red') +
  geom_line(aes(y = lower), linetype = 2, color = 'red') +
  geom_line(aes(y = upper), linetype = 2, color = 'red') +
  theme_bw())
}

map2(fits$data, fits$preds, bind_cols) %>%
  walk(plot_results)
```

## Crop Model Outputs

Import the results of the DSSAT simulations. Create a raster brick.
```{r}
yields_rast <- list.files('data/paleoD_10nov16/paleoD3p1_monthlyyieldonly/whemmer', 
                      pattern = '*ensemble*',
                      full.names = T) %>%
  map(raster) %>%
  brick %>%
  reclassify(c(-Inf, 0, 0, NA, NA, 0)) %>%
  mask(., land_mask, maskvalue = 0) %>%
  `names<-`(month.name) %>%
  `projection<-`('+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0')
```

Convert rasters to a data frame, adding in seasonal information.
```{r}
yields <- yields_rast %>% as.data.frame(xy = T, na.rm = T) %>%
  gather(month, yield, 3:14) %>%
  mutate(month = factor(month, levels = month.name),
         season = case_when(month %in% month.name[c(12, 1, 2)] ~ 'Winter', 
                            month %in% month.name[3:5] ~ 'Spring',
                            month %in% month.name[6:8] ~ 'Summer',
                            month %in% month.name[9:11] ~ 'Fall'),
         season = factor(season, levels = c('Winter', 'Spring', 'Summer', 'Fall')))
```

```{r eval=F, echo = F}
ggplot(yields, aes(x, y, fill = yield)) +
  geom_raster() +
  scale_fill_viridis() +
  facet_wrap(~month) +
  coord_quickmap() +
  ggtitle('Mean wheat yields by planting month') +
  theme_void()
```

```{r echo = F}
yields %>%
  group_by(x, y, season) %>%
  filter(yield == max(yield)) %>%
ggplot(aes(x, y)) +
  geom_raster(aes(fill = yield)) +
  scale_fill_viridis() +
  facet_wrap(~season) +
  ggtitle('Maximum wheat yields by planting season') +
  coord_quickmap() +
  theme_void()
```


```{r echo = F}
yields %>%
  group_by(x, y) %>%
  filter(yield == max(yield)) %>%
  filter(yield > 0) %>%
ggplot(aes(x, y)) +
  geom_raster(aes(fill = season, alpha = yield)) +
  geom_point(data = dates, aes(Long, Lat)) +
  geom_path(data = med_region_gg, aes(long,lat, group = group)) +
  scale_fill_manual(values = c('#2b83ba', '#abdda4', '#d7191c', '#fdae61')) +
  coord_quickmap(xlim = c(-9.9, 45), ylim = c(30, 50)) +
  ggtitle('Best planting seasons') +
  theme_void()
```

```{r eval = F}
vid <- ggplot(dat, aes(Long, Lat)) +
  geom_raster(data = best.month, aes(x, y, fill = Month, alpha = Yield)) +
  geom_point(aes(frame = Cal.BC * -1, cumulative = T)) +
    geom_path(data = med_region_gg, aes(long,lat, group = group)) +
  scale_fill_manual(values = c('#2b83ba', '#abdda4', '#d7191c', '#fdae61')) +
  coord_quickmap(xlim = c(-9.9, 45), ylim = c(30, 50)) +
  labs(title = 'Neolithic spread', subtitle = 'Initial site occupations over wheat planting season and yield') +
  theme_void()

gganimate(vid, filename = 'neo_spread.mp4', saver = 'mp4')
```

### Crop model costs
Now use the yield data to calculate new conductance matrices that reflect variable crop yields. Define a function that represents the relative costs of moving from a cell with optimal plantint date in one month to a a cell with another optimal planting date. Use the "weight" parameter to determine the steepness of this function.

```{r}
month_fun <- function(x){
  dist <- 6 - abs(abs(x[1] - x[2]) - 6) # distance in months
  return(exp(- 1 * dist^2))
}
```

```{r echo = F}
expand.grid(difference = seq(0,6,.1), weight = seq(.1,1,.1)) %>%
  mutate(test = exp(- weight * difference^2)) %>%
  ggplot(aes(difference, test, group = weight, color = weight)) +
  labs(x = 'Difference in optimal planting months', y = 'Conductance',
       title = 'Functions') +
  geom_line() +
    theme_minimal()
```
Use this function to calculate a transitionlayer.
```{r}
con_month <- which.max(yields_rast) %>%
  transition(month_fun, direction = 16)
```

## Cost Distances


Calculate the least cost distances from Jericho for each of the conductance layers, correcting for intercell distances.
```{r}
distances <- c(con_landsea,
               con_land,
               con_month,
               map(coastal_decay, ~. + con_land),
               map(coastal_decay, ~. + con_month)) %>%
  map(~ . * correction_lcp) %>%
  map(accCost, fromCoords = jericho) %>%
  brick
```

```{r echo = F}
distances %>%
  mask(land_mask, maskvalue = 0) %>%
  as.data.frame(xy = T, na.rm =T) %>%
  #rename(euc = layer.1, land = layer.2, landsea = layer.3, month_only = layer.4, month_sea = layer.5) %>%
  gather(type, cost, 3:13) %>%
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = cost)) +
  facet_wrap(~type, nrow = 3) +
  scale_fill_viridis(direction = -1) +
  coord_quickmap() +
  theme_void()
```


Add the results to the sites data. (Why does raster:extract result in some na's and infs? something to do with the land sea boundary?)
```{r}
costs <- dates %>%
  select(Long:Lat) %>%
  as.matrix %>%
  raster::extract(distances, ., df = T) %>%
  #rename(euc = layer.1, land = layer.2, landsea = layer.3, month_land = layer.4, month_sea = layer.5) %>%
  cbind(dates, .) %>%
  select(-ID) %>%
  gather(type, cost, 6:16) %>%
  nest(-type) %>%
  mutate(data = map(data, ~filter(., is.finite(cost)))) %>%
  mutate(data = map(data, ~filter(., cost < 1e10))) # shouldnt have to do this!
```


```{r}
quantile_gam <- function(df){
  qgam(list(
    Cal.BC ~ s(cost, bs = 'cr'), 
           ~ s(cost, bs = 'cr')), 
    data = df,
    qu = 0.85)
}

make_preds <- function(x){
  predict(x, type = 'response', se = T) %>%
  map(~.[,1]) %>%
  bind_cols %>%
  mutate(upper = fit + 2 * se.fit,
         lower = fit - 2 * se.fit)
}
```

```{r}
test <- costs$data[[1]]
plot(qgam(Cal.BC ~ s(cost, bs = 'cr') + te(Long, Lat),
     data = test,
     qu = .95))
```

```{r}
fits <- costs %>%
  mutate(mod = map(data, quantile_gam),
         aic = map_dbl(mod, 'aic'),
         preds = map(mod, make_preds))
fits
```


```{r}
plot_results <- function(x){
  print(ggplot(x, aes(cost, Cal.BC)) +
  geom_pointrange(aes(ymin = Cal.BC - Cal.BC.SD, 
                      ymax = Cal.BC + Cal.BC.SD,  
                      alpha = (1 / Cal.BC.SD) ^ 2), fatten = 2) +
  geom_line(aes(y = fit), color = 'red') +
  geom_line(aes(y = lower), linetype = 2, color = 'red') +
  geom_line(aes(y = upper), linetype = 2, color = 'red') +
  xlim(0, 5.5e6) +
  theme_bw())
}


map2(fits$data, fits$preds, bind_cols) %>%
  walk(plot_results)
```

```{r eval=FALSE, include=FALSE}
mod <- gam(Cal.BC ~ s(cost, bs = 'cr'), 
            data = costs$data[[3]],
            weights = (1/costs$data[[3]]$Cal.BC.SD)^2)
mod2<- gam(Cal.BC ~ s(cost, bs = 'cr'), 
            data = costs$data[[3]])
mod3 <- gam(Cal.BC ~ s(cost, bs = 'cr'), 
            data = costs$data[[3]],
            weights = (1/costs$data[[3]]$Cal.BC.SD)^2 / mean((1/costs$data[[3]]$Cal.BC.SD)^2))
plot(mod);plot(mod2); plot(mod3)
```

```{r}
test <- cbind(costs$data[[5]], pred = fits$preds[[5]]$fit) %>%
  mutate(resid = Cal.BC - pred) %>%
  filter(resid > 0)

yields %>%
  group_by(x, y) %>%
  filter(yield == max(yield)) %>%
  filter(yield >= 0) %>%
ggplot(aes(x, y)) +
  geom_raster(aes(fill = season, alpha = yield)) +
  geom_point(data = test, aes(Long, Lat, color = resid), size = 3) +
  geom_path(data = med_region_gg, aes(long,lat, group = group)) +
  scale_color_distiller(palette = 'RdYlBu') +
  coord_quickmap(xlim = c(-9.9, 45), ylim = c(30, 50)) +
  scale_fill_manual(values = c('#2b83ba', '#abdda4', '#d7191c', '#fdae61')) +
  labs(title = 'Neolithic spread', subtitle = 'Initial site occupation') +
  theme_void()

which.max(yields_rast) %>%
  as.data.frame(xy = T, na.rm = T) %>%
ggplot(aes(x, y)) +
  geom_raster(aes(fill = as.factor(layer))) +
  geom_point(data = test, aes(Long, Lat, color = resid), size = 3) +
  geom_path(data = med_region_gg, aes(long,lat, group = group)) +
  scale_color_distiller(palette = 'RdYlBu') +
  coord_quickmap(xlim = c(-9.9, 45), ylim = c(30, 50)) +
  labs(title = 'Neolithic spread', subtitle = 'Initial site occupation') +
  theme_void()
```



```{r eval = F}
library(tidygraph)
library(ggraph)

commuteDistance(con_landsea_r, pts)
net <- as_tbl_graph(out) %>%
  mutate(date = dat$Cal.BC, x = dat$Long, y = dat$Lat, resids = mod1$residuals) %>%
  activate(edges) %>%
  mutate(time_diff = .N()$date[from] - .N()$date[to]) %>%
  mutate(med_date = (.N()$date[from] + .N()$date[to]) / 2)


net %>%
  filter(time_diff > 0 & time_diff <= 65) %>%
  filter(weight <= 600000) %>%
ggraph() +
  geom_edge_fan(alpha = .2) +
  geom_node_point(aes(color = date)) +
  scale_color_distiller(palette = 'Spectral', direction = -1) +
  theme_void() +
  geom_path(data = med_region_gg, aes(long,lat, group = group)) +
  coord_quickmap(xlim = c(-9.9, 45), ylim = c(30, 50))
```

```{r eval = F}
dat2 %>%
  mutate(arrival = if_else(residuals <= 0, 'Late', 'Early')) %>%
ggplot(aes(Long, Lat)) +
  geom_raster(data = best.month, aes(x, y, fill = Month, alpha = Yield)) +
  geom_point(aes(color = arrival)) +
    geom_path(data = med_region_gg, aes(long,lat, group = group)) +
  scale_fill_manual(values = c('#2b83ba', '#abdda4', '#d7191c', '#fdae61')) +
  coord_quickmap(xlim = c(-9.9, 45), ylim = c(30, 50)) +
  labs(title = 'Neolithic spread', subtitle = 'Initial site occupations over wheat planting season and yield') +
  theme_void()
```