---
title: "EOF downscaling of Mid Holocene climate"
author: "Nick Gauthier"
date: "3/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(raster)
library(tidyverse)
library(zoo)
library(stars)
library(units)
library(tidyEOF)
sf_use_s2(FALSE)
```


# Setup

Setup the bounding box and a reference 5' raster for reprojection.

```{r}
bbox <- extent(c(-10, 45, 30, 50))

ref <- st_as_stars(st_bbox(bbox, crs = 4326), dx = 0.0833333)
ref_sim <- st_as_stars(st_bbox(bbox, crs = 4326), dx = 1)

coast <- rnaturalearth::ne_coastline(scale = 'medium', returnclass = 'sf') %>% 
  st_crop(ref)
countries <- rnaturalearth::ne_countries(scale = 'large', returnclass = 'sf') %>% 
  st_crop(ref)
```

Import the observational data.
warping it directly looses the unit conversion info from the geotiff file!
```{r}
era_land <- read_stars('/Volumes/Data/GCM//adaptor.mars.internal-1622741123.0372913-20427-10-9fe07582-7b92-49f5-83ea-f5e620542b4b.nc') %>% 
st_set_crs(4326) %>%
  .['tp'] %>%
  .[,-1,,1:444] %>%
 # as('Raster') %>%
 # stackApply(1:12, mean) %>%
 # st_as_stars() %>%
  #st_set_dimensions('band', values = getZ(brick('~/Downloads/adaptor.mars.internal-1622741123.0372913-20427-10-9fe07582-7b92-49f5-83ea-f5e620542b4b.nc')[[-c()]]), names = 'time') %>%
st_set_dimensions(., 'time', values = as.numeric(zoo::as.yearmon(st_get_dimension_values(., 'time')))) %>%
  setNames('prec') %>%
  mutate(prec = set_units(prec * 30, mm))

ggplot() +
  geom_stars(data = era_land) +
  coord_quickmap() +facet_wrap(~time) +
  scale_fill_viridis_c(na.value = NA)
```

```{r}
 get_pcs(era_land) %>%
  get_eigenvalues() %>% 
  plot_scree(k = 4, kmax = 20) +
  scale_color_brewer(palette = 'Spectral')
```
now squareroot
```{r}
 get_pcs(era_land, scale = TRUE) %>%
  get_eigenvalues() %>% 
  plot_scree(k = 4, kmax = 20) +
  scale_color_brewer(palette = 'Spectral')
```
```{r}
era_patterns_land <- get_patterns(era_land, k = 8)

ggplot() +
  geom_stars(data = get_correlation(era_land, era_patterns_land)) +
  facet_wrap(~PC) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA) +
  #scale_fill_viridis_c() +
  coord_quickmap() +
  theme_void()
```
```{r}
era_patterns_land_r <- get_patterns(era_land, k = 7, rotate = TRUE)

ggplot() +
  geom_stars(data = get_correlation(era_land, era_patterns_land_r)) +
  facet_wrap(~PC) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA) +
  #scale_fill_viridis_c() +
  coord_quickmap() +
  theme_void()
```

```{r}
era_patterns_land_s <- get_patterns(era_land, k = 7, scale = TRUE)

ggplot() +
  geom_stars(data = get_correlation(era_land, era_patterns_land_s)) +
  facet_wrap(~PC) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA) +
  #scale_fill_viridis_c() +
  coord_quickmap() +
  theme_void()
```


```{r}
era_patterns_land_rs <- get_patterns(era_land, k = 9, rotate = TRUE, scale = TRUE)

ggplot() +
  geom_stars(data = get_correlation(era_land, era_patterns_land_rs)) +
  facet_wrap(~PC) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA) +
  #scale_fill_viridis_c() +
  coord_quickmap() +
  theme_void()
```

```{r}
plot_amps(era_patterns_land_rs)
```

#seasonal decomp
```{r}
era_patterns_land_rs$amplitudes %>%
  mutate(month = as.numeric(factor(months(as.yearmon(time)), levels = month.name))) %>%
  mgcv::gam(PC1 ~ s(time) + s(month, bs = 'cc'), data = .)  %>%
  plot
era_patterns_land_rs$amplitudes %>%
  mutate(month = as.numeric(factor(months(as.yearmon(time)), levels = month.name))) %>%
  mgcv::gam(PC1 ~ te(time, month, bs = c('cr','cc')), data = .)  %>%
  plot(scheme= 1)
```

```{r}
era_patterns_land_rs$amplitudes %>%
  mutate(month = as.numeric(factor(months(as.yearmon(time)), levels = month.name))) %>%
  mgcv::gam(PC3 ~ s(time) + s(month, bs = 'cc', k = 12), data = .)  %>%
  plot
```
```{r}
era_patterns_land_rs$amplitudes %>%
  mutate(month = as.numeric(factor(months(as.yearmon(time)), levels = month.name))) %>%
  mgcv::gam(PC5 ~ s(time) + s(month, bs = 'cc'), data = .)  %>%
  plot
era_patterns_land_rs$amplitudes %>%
  mutate(month = as.numeric(factor(months(as.yearmon(time)), levels = month.name))) %>%
  mgcv::gam(PC5 ~ te(time, month, bs = c('cr','cc')), data = .)  %>%
  plot(scheme= 1)
```

```{r}
era_patterns_land_rs$amplitudes %>%
  mutate(month = as.numeric(factor(months(as.yearmon(time)), levels = month.name))) %>%
  mgcv::gam(PC5 ~ te(time, month, bs = c('cr','cc')), data = .)  %>%
  plot(scheme= 1)
```

autocorrelation at lag 12
```{r}
m1 <- era_patterns_land_rs$amplitudes %>%
  mutate(month = as.numeric(factor(months(as.yearmon(time)), levels = month.name))) %>%
  mgcv::gamm(PC1 ~ te(time, month, bs = c('cr','cc')), data = .) 

layout(matrix(1:2, ncol = 2))
 acf(resid(m1$lme), lag.max = 36, main = "ACF")
 pacf(resid(m1$lme), lag.max = 36, main = "pACF")
 layout(1)
```
```{r}
era_patterns_land_rs$amplitudes %>%
  pivot_longer(-time) %>%
  mutate(month = as.numeric(factor(months(as.yearmon(time)), levels = month.name)),
         year = as.numeric(floor(as.yearmon(time)))) %>%
  ggplot(aes(month,value,color = name)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_line(alpha = .1, aes(group = year)) +
  facet_wrap(~name, scales = 'free_y') +
    geom_smooth(formula = y ~  s(x, bs = 'cc', k = 12), method = 'gam') +
  scale_x_continuous(breaks = 1:12) +
  theme_bw()
```

```{r}
era_land_t <- read_stars('/Volumes/Data/GCM/adaptor.mars.internal-1622741123.0372913-20427-10-9fe07582-7b92-49f5-83ea-f5e620542b4b.nc') %>% 
st_set_crs(4326) %>%
  .['t2m'] %>%
  .[,-1,,1:444] %>%
st_set_dimensions(., 'time', values = as.numeric(zoo::as.yearmon(st_get_dimension_values(., 'time')))) %>%
  setNames('t2m') %>%
  mutate(t2m = set_units(t2m, K))
```

```{r}
 get_pcs(era_land_t, scale = TRUE) %>%
  get_eigenvalues() %>% 
  plot_scree(k = 4, kmax = 20) +
  scale_color_brewer(palette = 'Spectral')
```

```{r}
plot(era_land_t[,,,1:24])
```
```{r}
era_t2m <- get_patterns(era_land_t, k = 3, rotate = TRUE)

ggplot() +
  geom_stars(data = get_correlation(era_land_t, era_t2m)) +
  facet_wrap(~PC) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA, limits = c(-1, 1)) +
  #scale_fill_viridis_c() +
  coord_quickmap() +
  theme_void()
```
```{r}
plot_amps(era_t2m) + facet_wrap(~PC, nrow = 3)
```
```{r}
era_t2m$amplitudes %>%
  pivot_longer(-time) %>%
  mutate(month = as.numeric(factor(months(as.yearmon(time)), levels = month.name)),
         year = as.numeric(floor(as.yearmon(time)))) %>%
  ggplot(aes(month,value,color = year)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_line(alpha = .25, aes(group = year)) +
  facet_wrap(~name, scales = 'free_y') +
  scale_color_viridis_c() +
  geom_smooth(formula = y ~  s(x, bs = 'cc', k = 12), color = 'black', method = 'gam') +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_bw()
```

```{r}
ggplot() +
  geom_stars(data = st_apply(era_land_t, 1:2, mean)) +
  scale_fill_viridis_c(option = 'magma', na.value = NA) +
  theme_bw() +
  coord_quickmap()
```
```{r}
ggplot() +
  geom_stars(data = st_apply(era_land_t, 1:2, sd)) +
  scale_fill_viridis_c(option = 'magma', na.value = NA) +
  theme_bw() +
  coord_quickmap()
```

## CCSM
```{r}
mh <- brick('/Volumes/Data/PMIP3/MH/pr/pr_Aclim_CCSM4_midHolocene_r1i1p1_100001-130012-clim.nc', var = 'pr') %>%
  raster::rotate() %>%
  raster::crop(bbox, snap = 'out') %>%
  st_as_stars() %>%
  st_warp(ref_sim, use_gdal = TRUE, method = 'bilinear') %>%
  st_set_dimensions('band', values = 1:12, names = 'time') %>%
  setNames('prec') %>% 
  mutate(prec = units::set_units(prec, mm/s)) %>%
  mutate(prec = units::set_units(prec, mm/month))
```
```{r}
prec_ccsm <- brick('/Volumes/Data/GCM/pr_Amon_CCSM4_historical_r1i2p1_185001-200512.nc') %>%
    .[[1465:1872]] %>% # 34 years
   rotate() %>%
  crop(bbox, snap = 'out') %>%
 # stackApply(1:12, mean) %>%
  st_as_stars() %>%
  st_warp(ref_sim, use_gdal = TRUE, method = 'bilinear') %>%
  setNames('prec') %>%
    st_set_dimensions('band', values = 1:408, names = 'time') %>%
  mutate(prec = set_units(prec, mm/s) %>% set_units(mm/month))
```

```{r}
 get_pcs(prec_ccsm, scale = TRUE) %>%
  get_eigenvalues() %>% 
  plot_scree(k = 4, kmax = 20) +
  scale_color_brewer(palette = 'Spectral')
 get_pcs(era_pr_mon, scale = TRUE) %>%
  get_eigenvalues() %>% 
  plot_scree(k = 4, kmax = 20) +
  scale_color_brewer(palette = 'Spectral')
```
```{r}

```


```{r}
plot(mh)
```
## ERA
```{r}
ecmwf_cp_mon <- brick('data/ERAinterim_prec.nc', var ='cp') %>%
  rotate %>%
  crop(bbox, snap = 'out')# %>%
 # disaggregate(fact = 4, method = 'bilinear')

ecmwf_lsp_mon <- brick('data/ERAinterim_prec.nc', var ='lsp') %>%
  rotate %>%
  crop(bbox, snap = 'out') #%>%
  #disaggregate(fact = 4, method = 'bilinear')

era_pr_mon <- st_as_stars(ecmwf_cp_mon + ecmwf_lsp_mon) %>%
 # st_set_dimensions('band', values = getZ(ecmwf_cp_mon) %>% zoo::as.yearmon() %>% as.numeric(), names = 'time') %>%
  st_warp(ref_sim, use_gdal = TRUE, method = 'bilinear') %>%
  st_set_dimensions('band', values = as.numeric(zoo::as.yearmon(getZ(brick('data/ERAinterim_prec.nc', var ='lsp')))), names = 'time') %>%
  setNames('prec') %>%
  mutate(prec = set_units(prec, m/day) %>% 
           set_units(mm/month)) %>%
  .[,,,25:468]
```
```{r}
plot(era_pr_mon[,,,1:12])
```

```{r}
 get_pcs(era_pr_mon, scale = TRUE) %>%
  get_eigenvalues() %>% 
  plot_scree(k = 4, kmax = 20) +
  scale_color_brewer(palette = 'Spectral')
```

```{r}
era_patterns_mon <- get_patterns(era_pr_mon, k = 9, rotate = TRUE, scale = TRUE)
ggplot() +
  geom_stars(data = get_correlation(era_pr_mon, era_patterns_mon)) +
  facet_wrap(~PC) +
  scale_fill_distiller(palette = 'RdBu') +
  #scale_fill_viridis_c() +
  coord_quickmap()
```

```{r}
ccsm_patterns_mon <- get_patterns(prec_ccsm, k = 9, rotate = TRUE, scale = TRUE)
ggplot() +
  geom_stars(data = get_correlation(prec_ccsm, ccsm_patterns_mon)) +
  facet_wrap(~PC) +
  scale_fill_distiller(palette = 'RdBu') +
  #scale_fill_viridis_c() +
  coord_quickmap()
```

```{r}
ccsm_patterns_mon$eofs;era_patterns_mon$eofs
```

```{r}
congruence <- function(x, y) {
  # could check that both have the save dimensions
  t1 <- as_tibble(x$eofs) %>%
    pivot_wider(names_from = PC, values_from = weight) %>%
    select(-x, -y) %>%
    remove_missing() # not ideal but . . .
  
  t2 <- as_tibble(y$eofs) %>%
    pivot_wider(names_from = PC, values_from = weight) %>%
    select(-x, -y) %>%
    remove_missing()
  
  psych::factor.congruence(t1, t2) 
}

align <- function(x, y) {
  t1 <- as_tibble(x$eofs) %>%
    pivot_wider(names_from = PC, values_from = weight) %>%
    select(-x, -y) %>%
    remove_missing()
  
  t2 <- as_tibble(y$eofs) %>%
    pivot_wider(names_from = PC, values_from = weight) %>%
    select(-x, -y) %>%
    remove_missing()
  
  vegan::procrustes(as.matrix(t1), as.matrix(t2), scale = FALSE)

  #psych::factor.congruence(t1, t2) 
}
test <- align(ccsm_patterns_mon, era_patterns_mon)
summary(test)
test$rotation
test$Yrot
```

```{r}
era_patterns_mon$eofs[,,,c(2,6, 8, 7, 5, 3, 4, 9, 1)] %>% plot
```


```{r}
congruence(ccsm_patterns_mon, era_patterns_mon) %>%
    as_tibble(rownames = 'ref') %>%
  pivot_longer(-ref) %>%
  group_by(ref) %>%
  arrange(-abs(value), .by_group = TRUE) %>%
  mutate(fit = case_when(abs(value) >= 0.98 ~ 'Excellent',
                         abs(value) >= 0.92 ~ 'Good',
                         abs(value) >= 0.82 ~ 'Borderline',
                         abs(value) >= 0.68 ~ 'Poor',
                         TRUE ~ 'Terrible')) %>%
  filter(abs(value) == max(abs(value)))
```

```{r}
ggplot() +
  geom_stars(data = ccsm_patterns_mon$eofs) +
  facet_wrap(~PC) +
  scale_fill_distiller(palette = 'RdBu') +
  #scale_fill_viridis_c() +
  coord_quickmap()
ggplot() +
    geom_stars(data = era_patterns_mon$eofs[,,,c(2,6, 8, 7, 5, 3, 4, 9, 1)] %>%
                 split %>% mutate(PC7 = PC7 * -1, PC1 = PC1 * -1) %>%
                 merge(name = 'PC')) +
  facet_wrap(~PC) +
  scale_fill_distiller(palette = 'RdBu') +
  #scale_fill_viridis_c() +
  coord_quickmap()
```

```{r}
#need mid holocene monthly
mh_mon <- brick('/Volumes/Data/GCM/b40.mh6ka.1deg.003.cam2.h0.PRECT.080101-131612.nc', var = 'PRECT') %>%
  raster::rotate() %>%
  raster::crop(bbox, snap = 'out') %>%
  st_as_stars() %>%
  st_set_dimensions('band', names = 'time') %>%
  setNames('PRECT')%>% 
  mutate(PRECT = units::set_units(PRECT, m/s)) %>%
  mutate(PRECT = units::set_units(PRECT, mm/month))